---
title: "Trends in Time"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
#Load necessary packages
library(tidyverse)
library(ggridges)
library(plotly)
library(lubridate)

#Controlling figure output in markdown
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%",
  fig.align = "center",
  cache = FALSE
)

#Set Theme for ggplot2
theme_set(theme_bw() + 
            theme(plot.title = element_text(hjust = 0.5), 
                  legend.position = "bottom"))

#Set Scientific notation output for knitr
options(scipen = 999)

```

## Visualization

We look here at whether any noticable trends existed in time, if there are any seasonal effect, whether pit bulls bite less people over time, and whether borough-level effects exist. The figure below looks at the number of dog bites over time across all boroughs, broken down by breed type.  

```{r message = FALSE, warning = FALSE, error = FALSE, echo = FALSE}
#Read in dog bite data
dog_bite <- read_csv("./data/dog_bite_12.4.csv")

#Manipulate and tidy data for 
 time.overall.df <- dog_bite %>%
  mutate(date = ymd(date_of_bite)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  select(-date) %>% 
  mutate(borough = as.factor(borough)) %>% 
  group_by(year, month) %>%
  summarise(pitbull = sum(pit_bull),
            no_pitbull = sum(!(pit_bull))) %>% 
  ungroup() %>%
  gather(key = breed, value = `number of bites`, pitbull:no_pitbull) %>%
  mutate(
    breed = ifelse(breed == "pitbull", "Pit Bull", "Other") %>% as.factor(),
    date = year + (month - 1)/12
  )

# plot in ggplot
time.overall.plot <- time.overall.df %>%
 ggplot(aes(x = date, y = `number of bites`, colour = breed)) + 
  geom_point(alpha = 0.3, size = 2) + 
  geom_line(size = 1, alpha = 0.5) + 
  geom_smooth(size = 1.5, alpha = 0.5, se = F, method = "lm") + 
  theme(axis.text.y = element_text(color = "black", 
                                   size = 10,  hjust = 1), 
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1, size = 10)) +
  labs(
    x = "Time",
    y = "Number of Bites per Month",
    title = "Number of Bites Over Time"
  ) + 
    viridis::scale_colour_viridis(
    name = "Dog Breed", 
    discrete = TRUE
   ) +
  xlim(c(2015, 2018)) 

# print plot as plotly
ggplotly(time.overall.plot)
```

There is a moderate overall increase in the mean number of bites per month by non-pitbull breeds. However, for pit bulls, the trend in number of bites per month is more stable. In fact, there's evidence of a slight decrease in pit bull bites, on average. Additionally, fluctation is very evident within any given year, most likely corresponding to seasonal effects, with overall dog bites rising during the summer months and falling during the winter months. 

The figure below shows these bites trends by borough. In Manhattan, we see the largest mean difference in pit bull vs. non-pit bull bites per month; trends over time are similar between breed types. Conversely, a smaller mean difference between breeds was observed in Staten Island, with a nearly identical trends over time. Brooklyn and Queens both demonstrate a widening gap over time, with the pit bull bites decreasing and non-pit bull bites increasing notably. Interestingly, there is an inverse trend between pit bulls and other breeds; pit bull bites increase over time while non-pit bull bites decreased.

```{r message = FALSE, warning = FALSE, error = FALSE, echo = FALSE}
#Create Bites Per Month Data Frame, by Borough
time.boro.df <- dog_bite %>%
  mutate(date = ymd(date_of_bite)) %>% 
  mutate_at(vars(date), funs(year, month, day)) %>%
  select(-date) %>% 
  mutate(borough = as.factor(borough)) %>% 
  group_by(year, month, borough) %>%
  summarise(
    pitbull    = sum(pit_bull),
    no_pitbull = sum(!(pit_bull))
  ) %>% 
  ungroup() %>%
  gather(key = breed, value = `number of bites`, pitbull:no_pitbull) %>%
  mutate(
    breed = ifelse(breed == "pitbull", "Pit Bull", "Other") %>% as.factor(),
    date = year + (month - 1)/12
  )

#Lineplot of Bites per Month by Borough
time.boro.plot <- time.boro.df %>%
  filter(borough != "Other") %>%
  ggplot(aes(x = date, y = `number of bites`, colour = breed)) + 
  geom_line(size = 1, alpha = 0.6) + 
  geom_smooth(size = 1.5, alpha = 0.3, se = F, method = "lm") + 
  theme(axis.text.y = element_text(color = "black", 
                                   size = 10), 
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1, size = 10)) +
  labs(
    title = "Number of Bites by Borough over Time"
  ) + 
  viridis::scale_colour_viridis(
    name = "Dog Breed", 
    discrete = TRUE
   ) +
  xlim(c(2015, 2018)) +
  facet_grid(~borough)

#time.boro.plot
subplot(ggplotly(time.boro.plot)) %>% 
  layout(xaxis3 = list(title = "Time"), yaxis = list(title = "Number of Bites per Month"))
```


## Statistical Analysis

After visually inspecting the temporal dynamics of pit bull bites in time, we decided to formally test the hypothesis that the number of pit bull dog bites decreased from 2015-2018 in NYC. Motivated by our exploratory visualizations above, we also decided it was necessary to control for seasonal and borough-level effects. We fit a linear regression where our final model took the form

$$ Number \ of \ Bites_i \sim \beta_0 + \beta_1 Time_i + \beta_2 Pitbull_i + \beta_3 Pitbull_i:Time_i + \beta_4 Season_i + \beta_5 Borough_i + \varepsilon_i,$$
for the $i^{th}$ month. Results are below, and model assumptions are checked after the discussion.

```{r echo = FALSE}
# Data Frame for linear modeling
# Center time at 2015, borough factor, seasonal factor, filter 2015-2018
# Modeling with data grouped by month
lm.df <- time.boro.df %>%
  mutate(date = date - 2015,
         season = ifelse(month %in% c(12, 1, 2), "Winter", 
                         ifelse(month %in% c(3, 4, 5), "Spring", 
                                ifelse(month %in% c(6, 7, 8), "Summer",
                                       "Fall"))),
         season = as.factor(season),
         season = fct_relevel(season, "Winter", "Spring", "Summer", "Fall")) %>%
  rename(time = date) %>%
  filter(time >= 0 & borough != "Other")

# Full Model
lm.final <- lm(`number of bites` ~ time + breed + breed:time + season + borough, data = lm.df)

# Display
lm.final %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>%
  knitr::kable(digits = 3)
```

Most notably, our results indicate that for pit bulls we see an expected mean *decrease* of 1.91 bites per month for each additional month of observation, making the increase in time almost negligible after adjusting for seasonal and borough effects. Additionaly, we observed a baseline mean number of bites of 19.24 in the Bronx, in the winter, for non-pit bulls. From this baseline, we observed a moderate expected mean increase in bites per month overall (2.03, adj. breed, season, borough), and bites per month from the winter to the spring, summer, and fall respectively (6.89, 11.02, 5.49, adj. time, borough, breed). Compared to the Bronx, we also observed moderate expected bites per month increases in Brooklyn, Manhattan, Queens, and a decrease in Staten Island (5.95, 5.04, 9.39, -8.92, adj. time, season, breed). 

## Discussion

Ultimately, pit bull bites increased almost negligibly compared to other breeds (at the 10% significant level) in NYC from 2015-2018, adjusting for significant seasonal and borough level effects. Our initial intuitions that seasonal fluctuation and borough significantly affect the number of dog bites over time are also supported. Unfortunately, we cannot conclude whether media campaigns to destigmatize and rehabilitate pit bulls in recent years have directly caused the relative decline in pit bull aggression and citation which we observed. Still, we have identified a meaningful decreased trend in the amount of violent interactions with pit bulls in NYC, compared to other breeds.

## Appendix

Considering the assumptions for inference in our linear model, there is moderate deviance from normality in the upper tail of the Q-Q plot and a slight curvilinear trend in the residuals vs. fitted plot, but nothing too terrible. However, there are some outlying high count observations and the residuals vs. leverage plot (along with normality) are the most concerning barrier for inference.  

```{r echo = FALSE}
par(mfrow = c(2,2))
plot(lm.final)
```
